\subsubsection{Vice President,
 Finance}
\label{vice-president-finance}
The Vice President, Finance shall:

\begin{enumerate}
 \item
  Arrange for the custody and disbursement of funds of the MES, as per the MES Financial Policies (MES Bylaws Section \ref{financial-policies}).
 \item
  Be responsible for the MES financial accounts, both at a chartered bank or trust company and on campus (see MES Bylaws Section \ref{accounts-and-cash}).
 \item
  Ensure accurate and complete financial records are kept by employing data validation checkpoints.
 \item
  Prepare the operating budget of the MES before the Mid Summer Meeting of the year elected and continually update the ledgers in order to keep track of money spent and advise on further disbursement.
 \item
  Investigate and coordinate the investment of MES funds (see MES Bylaws Section \ref{investments}).
 \item
  Oversee development and/or signing of contracts including handbook printing companies and suppliers for Drain merchandise.
 \item
  Ensure that the President and VPSL have signing authority to charge fees to MES accounts.
 \item
  Oversee reimbursement processes from the special projects and student projects fund, in coordination with the Student Projects Coordinator(s).
 \item
 Organize, review, and approve funding requests for intramural sports teams (see MES Bylaws Section \ref{intramural-funding-policy})
 \item
  Be audited or reviewed by an outside accounting firm for the purpose of official statements if requested by Financial Services to receive Society fees.
 \item
  Serve on the macLAB Board of Directors (see MES Bylaws Section \ref{maclab-board-of-directors}).
 \item
  Arrange to have the combination lock for the MES safe changed when appropriate.
 \item
  Report to the President.
 \item
  Supervise the following Team positions:

  \begin{enumerate}
   \item
    Drain Coordinator(s)

  \end{enumerate}
\end{enumerate}