\section{Financial Policies}
\label{financial-policies}

\subsection{Budget}
\label{budget}
\begin{enumerate}
 \item
  The VPF should collect proposed budgets, financial accountability documents, and any other information at their discretion from groups during the month of June.
 \item
  The VPF should present the preliminary budget at the first Mid-Summer Council Meeting. This budget will contain the proposed accounts for the upcoming year.
 \item
  The VPF should make an effort to consult with the relevant stakeholders when preparing the budget.
 \item
  In the event of a non-mandatory MES membership fee, the VPF should make an effort to prepare for a variety of opt-out rates when preparing the budget.
 \item
  The VPF shall motion in the first meeting with the new council in April to approve the Executive Operations budget and any other funds to ensure that operations can continue throughout the summer.
 \item
  The VPF shall motion at a Council Meeting in Spring Term to approve the Welcome Week budget so that operations can continue throughout the summer. This cumulative budget shall not exceed the funds reserved for Financial Contingency.
 \item
  The MES Membership will approve the budget at a General Meeting in August; this approval authorizes the MES Executive to make necessary expenditures outlined in the budget without further approval.
 \item
  Changes to the approved budget:

  \begin{enumerate}
   \item
    The MES Council can amend the budget by vote during the year.
   \item
    Expenditures not outlined in the budget must be approved according to the MES Funding Policies (see MES Bylaws Section \ref{funding-policies})
   \item
    Budgetary changes will be moved via motion accompanied by the following:

    \begin{enumerate}
     \item
      Reason for the change.
     \item
      Why and which budget should be changed to accompany the proposed expenditure.
    \end{enumerate}
  \end{enumerate}
 \item
  Reserve funds shall be set according to MES Bylaws Section \ref{reserve-funds}.
 \item
  Contingency funds shall be set according to MES Bylaws Section \ref{contingency}.

\end{enumerate}

\subsection{Payment Policies}
\label{payment-policies}
\begin{enumerate}
 \item
  Documentation for approved expenses, including an expense report (Appendix E), must be submitted to the VPF by March 30th of the given term. Expenses from the month of April will be reported similarly before April 30th. % fix this lol - appendix
 \item
  Expenses of Program Societies, MES Groups, or MES Teams, and individuals qualifying and approved for reimbursement will receive their funding via electronic funds transfer (EFT) to their respective bank account or via cheques written for approved expense reports.
 \item
  At the discretion of the VPF, payment for an invoice shall be done for an approved expense request.
 \item
  The VPF shall ensure that all expenses include sufficient and honest documentation, in accordance with the funding request as mentioned in the Policy Manual.
 \item
  Payment for sponsorship funding requests may be dispensed at the discretion of the MES Executive.
 \item
  All reimbursements must be approved by a minimum of two signing authorities on the MES bank account. These authorities include the VPF, VPSL, and President. Approval may take the form of a signature on a cheque, or digital approval of an EFT payment.
 \item
  Gasoline/mileage reimbursement for eligible travel (determined by the VPF) expenses will be \$0.22 per kilometer. Supporting map data stating total shortest distance travelled must be submitted.
 \item
  Any MES Member may access MES financial records during VPF office hours.
 \item
  The MES credit card shall be used to make advance payments for large expenses or to make large online payments to avoid having MES members or affiliated parties float large amounts of personal money. ``Large'' is to be determined at the discretion of the VPF.

  \begin{enumerate}
   \item
    The VPF shall be the only one allowed to use the credit card to pay for online expenses. Regardless of how a purchase is made, the VPF shall be the only person to hold custody of the card and its payment information.
   \item
    The party wishing to have an expense charged to the MES Credit Card shall submit an explanation of what the item is for/how it will be used, and an Expense Report (Appendix E) for the purchase, along with an attached screenshot of the item and its price.
  \end{enumerate}
\end{enumerate}

\subsection{Funding Policies}
\label{funding-policies}

\subsubsection{Other Funding Requests}
\label{other-funding-requests}
All funding requests unlisted under the approved budget must be approved
by the appropriate level:

\begin{enumerate}
 \item
  Greater than \$1,500: General Meeting majority vote
 \item
  \$501 to \$1,500: MES Council majority vote
 \item
  \$500 or less: MES Executive majority vote

\end{enumerate}

\subsubsection{Societies Funding Policy}
\label{societies-funding-policy}
\begin{enumerate}
 \item
  A systematic method for the distribution of MES funds to all recognizedProgram Societies to encourage continuity for each society from year to year and to develop accountability to the MES for the funds awarded. % clerical
 \item
  Each Program society will receive a standard funding in the amount of \$500 per term. No Program Society will receive annual funding in excess of \$1000. Half of the annual funding allotment for each society is to be dispensed at the first General Meeting, with the remainder dispensed at the second General Meeting, except in the case of h) below. % clerical - randomly switching to lowercase s in Program Society, also below
 \item
  Program Societies failing to meet quorum requirements at either General Meeting will not receive the funding allotted for that meeting.
 \item
  Societies failing to meet quorum requirements at the first General Meeting will still be eligible to receive funding (only up to the first half of their annual allotment) based upon individual motions presented to and voted upon by the MES Council.
 \item
  Funds will be reimbursed as outlined under \ref{payment-policies}

\end{enumerate}

\subsubsection{Clubs and Teams Funding Policy}
\label{clubs-and-teams-funding-policy}
\begin{enumerate}
 \item
  A systematic method for the distribution of MES funds to all recognized MES Clubs and Teams, to avoid overlap with MSU funding, encourage continuity, and develop accountability to the MES, should be implemented.
 \item
  Each MES Club or Team is required to submit two packages to the DoC each year: a Budget Proposal Package and a Final Report Package.
 \item
  Unless otherwise written by the VPF, the Budget Proposal Package will be submitted by July 31st and it will include:

  \begin{enumerate}
   \item
    A budget proposal (following the template provided on the MES website or available from the VPF)
   \item
    A list and description of all proposed events for the coming year
   \item
    A statement of the MES Club or Team's starting balance (carry-over from the previous year)
  \end{enumerate}
 \item
  Unless otherwise determined by the VPF, the Final Report Package will be submitted at least 10 business days before the second General Meeting in a given academic year and it will contain:

  \begin{enumerate}
   \item
    An update of the original proposed budget
   \item
    A quantitative description of the distribution of MES Funds
   \item
    One event summary sheet for each event organized during the year
   \item
    A report of future recommendations and the overall functioning of the MES Club or Team (the VPF will use these figures to propose the following year's budget)
  \end{enumerate}
 \item
  The budget proposals will be assessed according to the following criteria:

  \begin{enumerate}
   \item
    Starting balance
   \item
    Type and variety of activities offered to students
   \item
    Level of activity
   \item
    Performance at the competitions they take part in
   \item
    Level and Sources of Sponsorship
   \item
    Level of fundraising activity
   \item
    Amount of funding requested from the MES
  \end{enumerate}
 \item
  The DoC shall review the budget proposals presented and suggest an amount to be awarded to each MES Club or Team. The MES Executives will review the suggested funding for each MES Club or Team and decide on the final amount that each club is to receive.
 \item
  All funding will be dispersed through approved expenses (see MES Bylaws Section \ref{payment-policies}). % wtf is this saying?
 \item
  If an MES Club or Team has not claimed any of their allotted funding by the beginning of second term, they must submit an activities report and progress update to the VPF by January 31st of the year, even if money is spent by the MES Club or Team during the month of January.

  \begin{enumerate}
   \item
    Once these documents are reviewed it is up to the VPF's discretion to either find the explanations satisfactory, or to arrange a meeting with the MES Club or Team to discuss their funding going forward.
   \item
    If these documents are not received, the MES Club or Team's funding will be reallocated to other MES Clubs and Teams.

  \end{enumerate}
\end{enumerate}

\subsubsection{Conference / Competition Funding Policy}
\label{conference-competition-funding-policy}
\begin{enumerate}
 \item
  Open conferences and competitions

  \begin{enumerate}
   \item
    The MES will fund up to 50\% of the delegate, transportation, and accommodation fees for all delegates.
   \item
    Delegations requesting funding from the Open Conference and Competitions Fund must fill out the MES Conference/Competition Application form (Appendix N) and submit it to the VPX.
   \item
    The funding shall be approved:

    \begin{enumerate}
     \item
      By a vote from the MES Council if in an amount greater than \$500, in which case the head delegate from the group requesting funding must present to the MES Council to receive the funding.
     \item
      At the discretion of the VPX if in an amount less than \$500, in which case the VPX must update the MES Council on what funding was approved.
    \end{enumerate}
   \item
    In the event that a presentation needs to be made, the presentation shall be made before attending the conference or competition and shall be evaluated based on the following criteria:

    \begin{enumerate}
     \item
      Delegate/entrance fees
     \item
      Location of the event
     \item
      Number of MES members to attend
     \item
      Length and reputation of the event
     \item
      The degree of effort put into fundraising by the delegates
     \item
      Benefits for the MES
     \item
      History and value of the conference or competition
    \end{enumerate}
   \item
    The head delegate must also submit a report to the VPX, to be published in The Frequency and/or the MES website no later than one month after the conference or competition has concluded.

    \begin{enumerate}
     \item
      The report shall include updates on activities of the delegates, tangible outcomes, and effectiveness of the delegation in representing McMaster University.
    \end{enumerate}
   \item
    The following funding mechanism will be used once the conference or competition has been approved for funding:

    \begin{itemize}
     \item
      Delegation is between 1-20 MES members -- 50\% of the delegate fees, transportation and accommodation up to a maximum for \$100 per delegate and 1 head delegate to a maximum of \$150, not exceeding \$1000 per delegation
     \item
      Delegation is between 20-50 MES members -- 50\% of delegate fees, transportation, and accommodation up to a maximum of \$75 per delegate and 2 head delegates to a maximum of \$150 each, not exceeding \$1500 per delegation
     \item
      Delegation is above 50 MES members -- 50\% of delegate fees, transportation \& accommodation up to a maximum of \$50 per delegate and 2 head delegates to a maximum of \$150 each, not exceeding \$2000 per delegation.
    \end{itemize}

   \item
    The delegation must submit the signed Delegate Code of Conduct (Appendix F) prior to attending the conference. Failing to do so will give the MES Executive the right to revoke reimbursement. % appendix
   \item
    MES funding approval (while not preferred) can be given post-event on the condition that the MES Executive has been made aware (in writing) of the event and that MES members are attending it. The amount given will still be determined by the above criteria.
   \item
    If the MES Council approves funding, the group must submit receipts for all expenses being covered by the MES no later than one month after the conference or competition has concluded.
   \item
    Inappropriate or unprofessional delegate behavior gives the MES Executives the right to revoke reimbursement.
   \item
    The funding pool is limited and may be exhausted before all delegations apply. The VPF shall decide whether requests are acceptable and within the limits of the MES Budget.
  \end{enumerate}

 \item
  Affiliate conferences
  \begin{enumerate}
   \item
    The MES will fund 100\% of the delegate, transportation, and accommodation fees for the VPX to attend all conferences that have a plenary

    \begin{enumerate}
     \item
      This funding is conditional and may be revoked if the VPX fails to give a presentation to the MES Council outlining the outcomes of the conference.
     \item
      The MES will fund 100\% of the delegate, transportation, and accommodation fees for up to two representatives (ideally the President and the VPX) to attend ESSCO Presidents' Meeting
    \end{enumerate}
   \item
    The MES will fund at least 50\% of the delegate, transportation, and accommodation fees for any other delegates requesting financial aid chosen by the Delegate Selection Committee.

    \begin{enumerate}
     \item
      All delegates must write an article or organize an event or activity, up to the discretion of the VPX.
     \item
      Delegates approved by the Delegate Selection Committee as a conference attendee but not for MES funding are able to attend the affiliate conference without any reimbursement of funds.
     \item
      It should be noted that if any delegate has already received funding for another affiliate conference in the same academic year, that that delegate can only be granted funding for an affiliate conference if the reimbursement rate remain consistent across all delegates other than the VPX and President for that same conference.
    \end{enumerate}
   \item
    The head delegate of the delegation receiving funding must fill out the MES Conference/Competition Application form (Appendix N) and submit it to the VPA.
   \item
    The delegation must submit the signed Delegate Code of Conduct (Appendix F) prior to attending the conference. Failing to do so will give the MES Executive the right to revoke reimbursement.
  \end{enumerate}
 \item
  Competitions

  \begin{enumerate}
   \item
    The MES will fund the delegate and transportation fees for McMaster Engineering Competition winners to attend the Ontario Engineering Competition, and Ontario Engineering Competition winners to attend the Canadian Engineering Competition.
   \item
    Winners of the McMaster Engineering Competition shall remit a deposit equivalent to 50\% of the delegate fee, with reimbursement contingent upon active participation in the Ontario Engineering Competition, both in the logistical planning preceding the event and during the event itself.
    \begin{enumerate}
     \item An exception may be granted on a financial need basis, at the discretion of the Vice President, External Relations.
    \end{enumerate}
   \item
    The McMaster Engineering Competition Chair(s) shall write an article for The Frequency and/or the MES Website to be published no later than two weeks after each competition, detailing the results of the competitions.
   \item
    The head delegate of the delegation receiving funding must fill out the MES Conference/Competition Application form (Appendix N) and submit it to the VPA.
   \item
    The delegation must submit the signed Delegate Code of Conduct (Appendix F) prior to attending the conference. Failing to do so will give the MES Executive the right to revoke reimbursement.

  \end{enumerate}
\end{enumerate}

\subsubsection{Intramural Funding Policy}
\label{intramural-funding-policy}
\begin{enumerate}
 \item
  The MES will sponsor intramural sports teams composed of at least 80\% MES members. These teams shall be reimbursed up to the percentage of MES members on their roster. If the intramural budget is underperforming the VPF and Sports Coordinator(s) reserve the authority to lower the mandatory member composition to a minimum of 75\% MES members. If this happens there would be retroactive reimbursements to teams from the first semester that had already submitted their funding documents.
 \item
  The MES will fund the cost of registration up to \$300, matching the percentage of MES members on the team, not including any required deposits.
 \item
  To receive funding, a team must submit the following items to the Sports Coordinator(s):

  \begin{enumerate}
   \item
    A completed copy of Appendix X -- Intramurals Funding, including a full list of the team's playoff roster, including programs of study and student numbers of each team member, signed by each member of the team, and indicating the percentage of MES members.
   \item
    A standard expense report indicating the value of funding which appropriately reflects the percentage of MES members and the deduction of any required deposits.
   \item
    Copy of the receipt as proof of registration.
  \end{enumerate}
 \item
  The Sports Coordinator(s) shall review requests for funding to ensure they meet all requirements, and forward expense reports and receipts to the VPF.
 \item
  The Sports Coordinator(s) will be notified of the budget by the VPF once the budget is approved. The Sports Coordinator(s) should only approve requests within the limits of the budget.
 \item
  The funding pool is limited and may be exhausted before all teams apply. The VPF shall decide whether requests are acceptable and within the limits of the MES Budget.
\end{enumerate}

\subsubsection{Special Projects Funding Policy}
\label{special-projects-funding-policy}
\begin{enumerate}
 \item
  MES members requesting funding for a special project must:

  \begin{enumerate}
   \item
    For funding requests less than \$500:

    \begin{enumerate}
     \item
      Submit the MES Special Projects Application form, created and maintained by the Student Projects Coordinator(s)
     \item
      Receive a majority vote from the VPSL, the VPF and the President
    \end{enumerate}
   \item
    For funding requests between \$500 and \$1000:

    \begin{enumerate}
     \item
      Contact the Administrator such that they will be on the agenda for an MES Council meeting.
     \item
      Submit a motion to be put forth at an MES Council meeting, stating the funding amount being requested and the project for which the funds are being requested.
     \item
      Present to the MES Council the purpose of the project, details of the funds requested, and the proposed benefit to students.
    \end{enumerate}
   \item
    For funding requests greater than \$1000:

    \begin{enumerate}
     \item
      Contact the Administrator such that they will be on the agenda for an MES General meeting.
     \item
      Submit a motion to be put forth at an MES General meeting, stating the funding amount being requested and the project for which the funds are being requested.
     \item
      Present to the MES Members the purpose of the project, details of the funds requested, and the proposed benefit to students
    \end{enumerate}
  \end{enumerate}
 \item
  MES Groups and Teams are not eligible for this funding.
 \item
  The funding pool is limited and may be exhausted before any application. The VPF shall decide whether requests are acceptable and within the limits of the MES Budget.

\end{enumerate}

\subsubsection{Student Projects Funding Policy}
\label{student-projects-funding-policy}
\begin{enumerate}
 \item
  MES members requesting funding for a student project must:

  \begin{enumerate}
   \item
    For funding requests less than \$500:

    \begin{enumerate}
     \item
      Submit the Student Projects Application form, created and maintained by the Student Projects Coordinator(s)
     \item
      Receive a majority vote from the VPSL, the VPF and the President
    \end{enumerate}
   \item
    For funding requests between \$500 and \$1000:

    \begin{enumerate}
     \item
      Contact the Administrator such that they will be on the agenda for a Council meeting.
     \item
      Submit a motion to be put forth at a Council meeting, stating the funding amount being requested and the project for which the funds are being requested.
     \item
      Present to Council the purpose of the project, details of the funds requested, and the proposed benefit to students.
    \end{enumerate}
   \item
    For funding requests greater than \$1000:

    \begin{enumerate}
     \item
      Contact the Administrator such that they will be on the agenda for an MES General meeting.
     \item
      Submit a motion to be put forth at a General meeting, stating the funding amount being requested and the project for which the funds are being requested.
     \item
      Present to the MES Members the purpose of the project, details of the funds requested, and the proposed benefit to students
    \end{enumerate}
  \end{enumerate}
 \item
  MES Groups and Teams are not eligible for this funding.
 \item
  The funding pool is limited and may be exhausted before any application. The VPF shall decide whether requests are acceptable and within the limits of the MES Budget.

\end{enumerate}

\subsection{Accounts and Cash}
\label{accounts-and-cash}

\subsubsection{Accounts}
\label{accounts}
\begin{enumerate}
 \item
  The MES shall be the owner of at least:

  \begin{enumerate}
   \item
    One account at an outside chartered bank or trust company
   \item
    One McMaster University account.
  \end{enumerate}
 \item
  The chartered bank account shall be used for the majority of daily operations such as deposits, purchases, and payment of bills.
 \item
  The VPF will administer the chartered bank account, which shall be a corporate account requiring three authorized signing officers, with two out of the three signatures required on each cheque. The three signing officers shall be the President, VPF and VPSL of the MES.
 \item
  The VPF is responsible for administering all cheques, transfers, and deposits, and must receive approval from the proper levels of the MES Council for all payments that have not been previously granted approval by way of the budget.

\end{enumerate}

\subsubsection{Investments}
\label{investments}
\begin{enumerate}
 \item
  Funds held by the MES during the school year, in particular in the period between receipt of student fees in October and payment of Engineering Co-op and Career Services (ECCS) fees in March, may be invested by the MES Executive in a higher return venture.
 \item
  The investment must be agreed upon unanimously by the MES Executive, and must involve no risk of loss of funds.
 \item
  The VPF and one of the other signing officers will carry out the investment.
 \item
  Investments will be handled through the chartered bank where the current MES external account is held, unless the MES Executive deems it necessary to use an alternate chartered bank or trust company.

\end{enumerate}

\subsubsection{Reserve Funds}
\label{reserve-funds}
\begin{enumerate}
 \item
  The MES shall maintain a flexible (no risk) investment account, the sum of which shall constitute a reserve surplus fund (ECCS Fees held by the MES do not count towards this surplus, see MES Bylaws Section \ref{investments}).
 \item
  Funds from this reserve shall be used at the discretion of the MES Executive as per Funding Policy (see MES Bylaws Section \ref{funding-policies}).
 \item
  MES members requesting funding from Reserve Funds must submit a proposal to the MES Executive.
 \item
  This fund shall not be used to finance any of the operations mentioned within the operating budget of the fiscal year. (Such as, but not limited to -- Social Events, Groups \& Teams, Conferences, etc.)
 \item
  For all approved projects, the individual/group must submit receipts for all expenses being covered by the MES prior to receiving reimbursement. The individual/group must also meet all requirements detailed in the Sponsorship Checklist (see Appendix AA) before receiving reimbursement.

\end{enumerate}

\subsubsection{Contingency}
\label{contingency}
\begin{enumerate}
 \item
  MES recognizes that prudent management of the society's resources require that the following funds be set aside in order to protect the MES in times of adverse economic condition or where the need for major expenditure may arise:

  \begin{enumerate}
   \item
    The Operating Contingency budget shall be set to \$30,000. The VPF shall allocate some of the Operating Contingency funds towards other operating budgets, as the year goes on and the operating finance risks are low.
   \item
    The Financial Contingency budget shall be set to \$70,000. Any expenditure from this budget shall be only done at the discretion of the President and the VPF, and should be replenished as soon as possible. This budget shall be treated as Mandatory Retained Earnings, and should be held separate from the MES Operating Budget.

  \end{enumerate}
\end{enumerate}

\subsubsection{Petty Cash}
\label{petty-cash}
\begin{enumerate}
 \item
  Petty cash may only be held by the VPF to facilitate the sale of tickets to social events.
 \item
  Petty cash will be held using the cash box or the safe in the MES Office. The cash box must be kept in a secure location at all times.
 \item
  Petty cash in the safe should not exceed \$5000.
 \item
  Large amounts of money that are to be deposited in the bank may be kept in the safe by the VPF for a short period of time pending the next bank deposit.
 \item
  The VPF should empty the safe at least once a month.
 \item
  The combination lock for the safe shall be changed once a year immediately after the VPF has been ratified into the position. Only the Vice President Finance and President shall know the combination.

\end{enumerate}

\subsection{Student Fees}
\label{student-fees}
\begin{enumerate}
 \item
  MES student fees are collected yearly with tuition from each undergraduate engineering student by McMaster Financial Services.
 \item
  The per student MES student fee must be represented explicitly in the budget posted to the MES website upon approval by the MES council.
 \item
  All of the funds created by these fees are turned over to the MES by Financial Services in October by way of a cheque. This cheque is to be deposited in the external account such that the money can be used to cover operating expenses during the school year, with the excess being invested in a low-risk venture through the bank.
 \item
  An additional \$50 voluntary contribution will be collected with MES fees from each undergraduate engineering student for the McMaster Laboratory Advancement Benefaction Endowment Fund.
 \item
  An additional \$50 fee will be collected with tuition from each undergraduate engineering student to pay off the construction of the Gerald Hatch Centre.
 \item
  Fees shall be increased by the Consumer Price Index upon a majority vote by the MES Executive.

\end{enumerate}

\subsection{Accounting}
\label{accounting}
\begin{enumerate}
 \item
  The VPF shall keep comprehensive records of all transactions through the MES accounts, by way of books that are to be kept in accordance with good accounting practice. MES shall employ a bookkeeper at the discretion of the VPF.
 \item
  Records may be kept through use of personal computer accounting software, which is to be available on the MES office computer.
 \item
  The VPF shall have the books balanced at the end of each semester and prepared for on demand perusal.
 \item
  The VPF reserves the right to transfer fees to a student account should a cheque not clear due to insufficient funds on part of the MES.

\end{enumerate}

\subsection{Endowment Funds}
\label{endowment-funds}
\subsubsection{The McMaster Laboratory Advancement Benefaction Endowment Fund (i.e. macLAB)}
\label{the-mcmaster-laboratory-advancement-benefaction-endowment-fund-i.e.-maclab}
\begin{enumerate}
 \item
  The administration of the McMaster Laboratory Advancement Benefaction Endowment Fund is the responsibility of the macLAB Board of Directors.
 \item
  The macLAB Board of Directors will administer the fund in accordance with the macLAB Bylaws.
 \item
  The macLAB Bylaws may be changed and updated without approval from the MES Council.

\end{enumerate}

\subsection{Honoraria \& Awards}
\label{honoraria-awards}
\begin{enumerate}
 \item
  The MES Executive may choose to honor any MES member by way of an award or gift.
 \item
  The purchase of these awards or gifts must have direct approval from the MES Council before the purchase is made.

\end{enumerate}

\subsection{Donations}
\label{donations}
\begin{enumerate}
 \item
  All donations made by the MES shall fall under normal financial policies for release of funds, based on dollar amount, except in the case where the money is raised by fundraising techniques.
 \item
  Proof of donation must be provided to the VPF for records.
 \item
  All donations made by the MES must be publicly reported.
\end{enumerate}

% \end{document}
